Collaborative Visual Analysis on the Web with RCloud

### 2015-06-18 casestudies.tex
try to clean up text

As an example, we present a simple real-world application
deployed under RCloud.

The application is an implementation of LDAVis,
a recent method for visualizing large text corpora.
It was designed to help non-experts to explore
collections of short text documents using topic modeling and
visualization. Topic modeling~\cite{Blei:2003:LDA} is a standard
technique for text summarization which, although powerful, requires
some manual intervention and interpretation~\cite{Sievert:2014:LAM}.

@@

LDAVis combines text analysis, dimensionality reduction and
interactive visualization. Text analysis is performed by
combining a standard R library to fit LDA models using
Gibbs sampling~\cite{} with custom R code written by the
developers. The analysis module is, ultimately, a single R function
exposed to the web application via the Javascript-R RPC
mechanism described in Section~\ref{sec:system};
thus analysis is performed remotely on an RCloud server.

@@

The LDAVis application is simple, but highlights some unique features
of RCloud. While RCloud notebooks allow deployment of R analyses over
the web with no additional effort, RCloud \emph{applications} are more
powerful, and are developed with a combination of Javascript and HTML
for the front end. This requires additional knowledge over Shiny, but
we argue that the RCloud model makes the analysis side simpler 
(since analysts simply write R in the style they are used to),
and the front end visualization side is simpler for the web
developers (since they simply write the Javascript code in
the style \emph{they} are used to). In addition, RCloud applications
inherit the automatic deployment and discoverability features of
all RCloud notebooks.

%% Text analysis in

%% To illustrate some aspects of RCloud, we present an example notebook
%% for stock price analysis.\footnote{Unfortunately this report could not
%% include production notebooks containing proprietary information.}
%% The example, although simplified, shows key steps in the development
%% and deployment of an application.

%% \begin{figure}
%% \includegraphics[width=\linewidth]{fig/casestudy1/casestudy1.pdf}
%% \caption{\label{fig:stockvis}Iterations of a stock ticker
%%   visualization, based on an example by Hadley Wickham. A
%%   simple, static visualization of the closing price of a single stock
%%   is progressively developed into a configurable display suitable
%%   for a dashboard, then into an interactive visualization that
%%   compares the volatility and volume of two stocks, and finally
%%   into an API for data access by other RCloud notebooks. The notebooks
%%   in this example can all be loaded as web pages. When a notebook
%%   corresponding to a function call is displayed a web page,
%%   its associated documentation is displayed.}
%% \end{figure}

%% It shows a sequence of visualizations of the performance
%% of financial stocks over several years. The first visualization
%% in this example uses ggplot2 and was written by Hadley Wickham.
%% It reads data provided by Yahoo! Finance as a web service, and shows
%% the price for a single trading symbol.

%% The webpage that produces that visualization includes a link to
%% the underlying source code. From this link, a user can fork the notebook.
%% In the example,a configurable ticker is added, based on the URL of
%% the notebook. The change to the notebook is minor.

%% Next, the notebook author or a collaborator may decide to extend
%% the application to also provide convenient access to the data,
%% apart from its visualization.
%% This is done by creating a notebook that defines a function.
%% That notebook then becomes a \emph{subroutine} for other notebooks.
%% It can be invoked from interactive notebooks, such as dashboards.
%% The data access notebook is version-controlled like other notebooks.

%% Finally, consider the situation where an analyst wants to understand
%% the price dynamics of stocks with respects to other attributes and
%% time ranges. For this, interactive visualization may be very helpful.
%% In our example, the analyst creates an interactive tool with multiple
%% linked views.

%% Demonstrating the integration of R and JavaScript, here the analyst
%% writes an R function that passes the dataframe and description of
%% the charts to the dcplot library.  Simple R expressions
%% are captured as trees to generate JavaScript expressions.
%% The terse chart description language, with sensible defaults inspired by
%% ggplot2, provides a simple yet powerful interface to the grouping
%% and reduction functionality of the well-accepted charting libraries
%% crossfilter, dc.js and d3.js.

% \subsection{Text analysis\label{sec:textvis}}
% Same.

>>>>

To illustrate ...
As an example, we present a simple real-world application
deployed under RCloud.
The application is an implementation of LDAVis,
a method for visualizing large text corpora.
LDAVis helps non-experts to explore collections of
short text documents using topic modeling and
visualization. Topic modeling~\cite{Blei:2003:LDA}, although
powerful, often requires human interaction
and interpretation~\cite{Sievert:2014:LAM}.

@@

LDAVis combines text analysis, dimensionality reduction and
interactive visualization. Text analysis is performed by
combining a standard R library to fit LDA models using
Gibbs sampling~\cite{} with custom R code written by the
developers. The analysis module is a single R function
exposed to the web application via the Javascript-R RPC
mechanism described in Section~\ref{sec:system};
thus analysis is performed remotely on an RCloud server.

@@

LDAVis highlights some unique features of RCloud.
While RCloud notebooks allow deployment of R analyses over
the web with no additional effort, RCloud \emph{applications} are more
powerful, and are developed with a combination of Javascript and HTML
for the front end. This requires additional knowledge over Shiny, but
the RCloud model makes the analysis side simpler (since analysts
simply write R in the style they already know)
and front end visualization side is simpler for the web
developers (since they simply write the Javascript code in
the style \emph{they} are used to). In addition, RCloud applications
inherit the automatic deployment and discoverability features of
all RCloud notebooks.

### 2015-06-18 discussion.tex
try to clean up text

\paragraph*{Security}
It is essential to provide security in an environment for
collaboration and data publishing. To provide some
protection for RCloud workbooks in an organization's intranet,
RCloud uses an object capability model
\cite{Miller:2006:RCT}
recently added to the Rserve protocol \cite{Urbanek:2003:AFW}
that prevents unauthenticated clients from making
unauthorized calls to the RCloud runtime environment.

Our back-end environment assumes a high degree of trust between users.
Access control for information security is delegated to the host
operating system and web server. In practice, most operating
systems and web servers rely either on coarse permission models,
which tend to be ineffective, or on detailed access control
lists, which tend to be cumbersome.
Information security in RCloud should be improved. More
sophisticated approaches, such as information flow analysis,
and query languages that ensure differential privacy, are
active research topics, and many problems remain
\cite{Moore:2011:SAF}. We are hopeful that having a richer
environment for collaboration and information sharing will encourage
new approaches to information security in visual analytics.
%\gordon{This sounds like we think we are helping by pushing the
%system to the breaking point, is that the idea?}

\paragraph*{Exploration is hard}
R is a popular language for exploratory data analysis, and we built RCloud
as a way for data scientists to seamlessly transition from EDA to sharing
and presentation. But the combined notebook and command line interface we
designed interferes with the use of R for exploration.

In existing R tools, the command prompt is used to try things out. It is very
fast and there is no commitment: nothing gets saved and there is nothing to
clean up. In contrast, in RCloud's notebook interface, the extra cells need
to be deleted later, and there seems to be a mental burden that \emph{bad stuff
is getting saved}.

One thing to try here is making deletion easier, for example with a
shortcut key to remove the last cell, or a way to keep a few cells and cull
the rest. A far more ambitious solution, suggested by one of our
interviewees, is to implement auditing of data analyses as it existed in S
\cite{Becker:1988:Auditing}, so that a result could be selected and code
which did not lead to it could be discarded automatically.

@@

\paragraph*{Discovery is hard}
While we provided tools for searching and curating notebooks, the
search function does not help for finding methodology (and does not protect
searchers from erroneous notebooks), and the notebook tree starts to get
unwieldy with hundreds of users and many hundreds of notebooks. In addition,
our choice to retain notebooks even when they have been deleted 
exacerbates the situation, since obsolete and broken notebooks stay around
forever.

One thing to try is tagging and filtering on tags, so that users
can mark their own notebooks or those of others to make them easier to
find. However, this relies on users making a conscious effort, which
is not always reliable. It seems better if possible to lean more on
passive metrics, such as how often a notebook has been run.

The system also needs to help users to curate notebooks. If a notebook
has been deleted or superseded by one of its forks, a user who opens it
should be warned. And operations on multiple notebooks and folders of
notebooks should be supported to make curation easier.

\paragraph*{Collaboration is hard}
While forking provides a simple way to continue someone else's work, working
in teams is a lot more complicated. The forking feature is popular but it
poses a problem, too, as dozens of versions of a notebook proliferate.

One solution we are considering is to allow overwriting the current state
of a notebook with the current state of another fork of the notebook.

A more general solutions would be a serious engineering endeavor. Even
GitHub's admirable web interface does not support merging where there are
conflicts. Unrestricted sharing of notebooks would probably require moving
toward git's notion of decentralized code repositories, because GitHub gists
do not support shared ownership.

>>>>

\paragraph*{Security}
Security is an important concern in an environment for
collaboration and data publishing. To provide some
protection for RCloud workbooks in an organization's intranet,
RCloud uses an object capability model
\cite{Miller:2006:RCT}
recently added to the Rserve protocol \cite{Urbanek:2003:AFW}
that prevents unauthenticated clients from making
unauthorized calls to the RCloud runtime environment.

Our back-end environment assumes a high degree of trust between users.
Access control for information security is delegated to the host
operating system and web server. In practice, most operating
systems and web servers rely either on coarse permission models,
which tend to be ineffective, or on detailed access control
lists, which tend to be cumbersome.
Information security in RCloud should be improved. More
sophisticated approaches, such as information flow analysis,
and query languages that ensure differential privacy, are
active research topics, and many problems remain
\cite{Moore:2011:SAF}.
% We are hopeful that having a richer
% environment for collaboration and information sharing will encourage
% new approaches to information security in visual analytics.
%\gordon{This sounds like we think we are helping by pushing the
%system to the breaking point, is that the idea?}

\paragraph*{Exploration is hard}
% R is a popular language for exploratory data analysis, and we built RCloud
% as a way for data scientists to seamlessly transition from EDA to sharing
% and presentation.
Users reported that RCloud's combined notebook and command line interface
interferes with the familiar use of R for exploration.
In current R practice, the command prompt is used to try things out.
It is very fast and there is often no commitment: nothing gets saved
and there is nothing to clean up.
In contrast, in RCloud's notebook interface, the extra cells need
to be deleted later, and there seems to be a mental burden
that \emph{scratch work is getting saved}.

One thing to try is to make deletion easier, for example with a
shortcut key that removes the last cell, or a way to keep a few cells
and cull the rest. A far more ambitious solution, suggested by one of our
interviewees, is to implement auditing of data analyses as it existed in S
\cite{Becker:1988:Auditing}, so that a result could be selected and code
which did not lead to it could be discarded automatically.

@@

\paragraph*{Discovery is hard} There are several problems.
With hundreds of users and thousands of notebooks, the notebook
tree becomes unwieldy.
The search function does not protect searchers from obsolete and
erroneous notebooks, which by default are retained forever.
It helps with finding notebooks, but not with acquiring
other types of information, such as methodology.

One thing to try is tagging, so that users can mark notebooks
to make them easier to find. However, this relies on users making
a conscious effort, which is not always reliable. It seems better
to rely more on passive metrics, such as how often a notebook has been run.

The system also needs to help users to curate notebooks. If a notebook
has been deleted or superseded by one of its forks, a user who opens it
should be informed. Operations on multiple notebooks and folders of
notebooks should be supported to make curation easier.

\paragraph*{Collaboration is hard}
While forking provides a simple way to continue someone else's work, working
in teams is a lot more complicated. The forking feature is popular but it
poses a problem, too, when dozens of versions of a notebook proliferate.
A possible solution is to allow overwriting the current state
of a notebook with the current state of another fork of the notebook.

Implementing a more general solution seems like a big job.
Even GitHub's web interface does not support merging where there are
conflicts. Unrestricted sharing of notebooks may involve moving
toward git's notion of decentralized code repositories, because GitHub
gists do not support shared ownership.

### 2015-06-18 interviews.tex
try to clean up text

Besides providing a way to present work, notebook sharing can provide a
starting point for coding. Lilo says, ``The best part is how easily you can
share code. You can find a working example, rather than wearing
out Google and finding questionable examples that may or may not work.'' Wendy
notes, ``[If] some person has done something similar, then you're able to just
edit that, and that's saved a lot of work time for me.''

@@

\subsection{Forking}
The ability to start work where someone else let off, by forking, proved to
be a popular feature. In fact, almost twice as many (131) users have forked
someone else's notebook as have starred one (75).  Lilo says, ``It's one of the
handiest functions, because instead of having to find it, copy, paste it, you
just hit Fork, rename it, and it's done. It's pretty amazing.''

@@

A common but somewhat problematic use of forking is to change
parameters. Wendy says, ``I've been forking other people's notebooks [because] I
want to run them on a different part of data, or I want to change some parts,
I don't want to see this column, I want a different column, things like
that.''  Eric complains that a notebook might say ```This is a report of
the volume of all of our feeds for this month', and someone would want to
look at it for the next month or the previous month, so they'd fork it to
change the month.''

Parameters can be added to a notebook URL, but adding user interface
elements to do the same thing requires expertise. Tool builders see a
need for facilities to make it easier. Allison calls it ``web-enabling'' the
notebook: ``[users] can always fork the notebooks and make changes, but I
feel that if the owner of the notebook web-enables it, it's easier to
run. Instead of forking it, if they can set options, it's probably more
efficient, and they can [still] fork it if they want to.''

@@

Iris points out that the automatic versioning works well for dealing with the
minutiae of web development:
\begin{quote}
I like the fact that it has a built-in editor, so
if you need to fix a typo in a link, or an extra line break or other
nonsense, you can just switch to the edit view, pull up your asset, type
something, it's automatically saved, committed, everything. You don't have to go
back to your source code, change it, commit it to the repo, pull the repo to
your distribution version. 
%% None of that's necessary. A lot of web development is
%% fixing stupid little bugs, and you can do that instantly without a lot of
%% overhead, which is nice.
\end{quote}

@@

More selective ways to search will be needed as the number of notebooks
grows further. We explore some ideas in Section~\ref{sec:discussion}.

@@

Although most of the analysts use and appreciate the sharing features,
RCloud is less popular as a tool for exploratory data analysis. Every
analyst with prior R experience still prefers to do analysis in another
tool, and paste code into RCloud for sharing.

This is mainly out of familiarity. Eric asks, ``Why would I want to use
RCloud over my current setup? If it's just me, I like my text editor and
terminal. There's nothing that I want that those two don't give me.''

@@

Coby also works with text files and command line, rearranging a file so the
good code is at the top of the file. When he's done, all the right code is
at the top. RCloud does not readily support this workflow, but Coby
often shares work by pasting it into an RCloud notebook when he's done.

@@

Working in a shared environment also entails compromises about what you can
install. Joy says installing alternative or nonstandard packages is
intrinsic to exploratory data analysis.

@@

Simply the difference between a cell and the command prompt can confuse some people.
Kenyon explains:
\begin{quote}
If I was typing into one of the cells near that top, I had to think ``I'm editing
this cell and then I have to execute it,'' and if I was editing one at the bottom,
I could type it but then it would automatically execute [by pressing enter].
%, then it became a standard cell and I have to edit it.
\end{quote}

@@

The cell structure also can be problematic if some cells take a long time to
execute. In Coby's work, there is often a ``long tail''.  The first cell may
take just seconds to run, the next cell minutes, and the last cell 5 hours.
In this situation, the Run Whole Notebook button is dangerous.
Coby reports that when converting his work to notebooks, he
ends up with a lot of comments saying ``This cell takes a long time to run.''

To avert this pitfall, some users write code within cells to explicitly cache results.
Coby ends up ``littering the notebook with little switches that comment out the
slow parts'' and either load or save the object to disk. One of the authors of this paper
manually writes cells that check if a result file exists, and perform a calculation only
if it's needed.

>>>>

Besides providing a way to present work, notebook sharing can provide a
starting point for coding. Lilo says, ``The best part is how easily you can
share code. You can find a working example, rather than wearing
out Google and finding questionable examples that may or may not work.'' Wendy
notes, ``[If] some person has done something similar, then you're able to just
edit that, and that's saved a lot of work for me.''

@@

\subsection{Forking}
The ability to start work where someone else let off, by forking, was
a popular feature. Almost twice as many (131) users have forked
someone else's notebook as have starred one (75).  Lilo says, ``It's one of the
handiest functions, because instead of having to find it, copy, paste it, you
just hit Fork, rename it, and it's done. It's pretty amazing.''

@@

A common but ill-advised use of forking is to change
parameters. Wendy says, ``I've been forking other people's notebooks [because] I
want to run them on a different part of data, or I want to change some parts,
I don't want to see this column, I want a different column, things like
that.''  Eric complains that a notebook might say ```This is a report of
the volume of all of our feeds for this month', and someone would want to
look at it for the next month or the previous month, so they'd fork it to
change the month.''

Parameters can be added to a notebook URL, but adding user interface
elements to do the same thing takes more expertise. Tool builders see a
need for facilities to make it easier. Allison calls it ``web-enabling'' the
notebook: ``[Users] can always fork the notebooks and make changes, but I
feel that if the owner of the notebook web-enables it, it's easier to
run. Instead of forking it, if they can set options, it's probably more
efficient, and they can [still] fork it if they want to.''

@@

Iris points out that automatic versioning works well for coping with
the details of web development:
\begin{quote}
I like the fact that it has a built-in editor, so
if you need to fix a typo in a link, or an extra line break or other
nonsense, you can just switch to the edit view, pull up your asset, type
something, it's automatically saved, committed, everything. You don't have to go
back to your source code, change it, commit it to the repo, pull the repo to
your distribution version. 
%% None of that's necessary. A lot of web development is
%% fixing stupid little bugs, and you can do that instantly without a lot of
%% overhead, which is nice.
\end{quote}

@@

More selective ways to search will be needed as the number of notebooks
grows further. We mention possible approaches in Section~\ref{sec:discussion}.

@@

Although most of the analysts appreciate RCloud's sharing features,
it is not popular as a tool for exploratory data analysis.
All analysts with R experience prefer to do analysis in another
tool, then paste code into RCloud for sharing.
It appears this is mainly due to familiarity.
Eric asks, ``Why would I want to use
RCloud over my current setup? If it's just me, I like my text editor and
terminal. There's nothing that I want that those two don't give me.''

@@

Coby also works with text files and commands, rearranging source files
so the good code is at the top. RCloud does not readily support this
workflow, so Coby often shares work by pasting it into an RCloud
notebook when he's done.

@@

Working in a shared environment also entails compromises about what you can
install. Joy says installing alternative or nonstandard packages is
intrinsic to exploratory data analysis. This is a weakness in systems
like RCloud.

@@

The difference between a cell and a command prompt confuses some people.
Kenyon explains:
\begin{quote}
If I was typing into one of the cells near that top, I had to think ``I'm editing
this cell and then I have to execute it,'' and if I was editing one at the bottom,
I could type it but then it would automatically execute [by pressing enter].
%, then it became a standard cell and I have to edit it.
\end{quote}

@@

The cell structure also can be problematic if some cells take a long time to
execute. High variance is common: cells may take anywhere from seconds to
hours to run.  In this situation, the Run Whole Notebook button is dangerous.
Coby reports that when converting his work to notebooks, he
ends up with a lot of comments saying ``This cell takes a long time to run.''

To avoid this pitfall, some users write code within cells to explicitly cache results.
Coby ends up ``littering the notebook with little switches that comment out the
slow parts'' and either load or save the object to disk. One of the authors of this study
manually writes cells that check if a result file exists, and perform a calculation only
if it's needed.
